---
title: "Analyze the output of metagenome-atlas"
output: html_document
---



# Installing tools

!! If a yellow banner appears that ask you to install libraries click on accept.

```{r install}
# packages that are not on CRAN

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

bioconductor_packages <- c("phyloseq", "microbiome", "ComplexHeatmap", "ggtree")
BiocManager::install(setdiff(bioconductor_packages, rownames(installed.packages())), update = F)



# microviz can be installed from this repository

if (!requireNamespace("microViz", quietly = TRUE)) {
  install.packages("microViz", repos = c(
    microviz = "https://david-barnett.r-universe.dev/",
    CRAN = "https://cloud.r-project.org"
  ))
}
```

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE, include = TRUE
)
```

Load libraries

```{r, libraries,include=FALSE}
# library(plotly) # don't load due to import conflicts
library(heatmaply)
library(dplyr) # dpyr masks select from plotly
library(readr)
library(stringr)
library(tidyr)
library(tibble)
library(ggplot2)
library(ggbeeswarm)
library(pheatmap)
library(grid)
library(microbiome)
library(ape)
library(vegan)
library(useful)
library(kableExtra)
library(ggtree)

library(arrow)
library(yaml)
library(phyloseq)
library(tidytree) # overwrites tree from phyloseq


library(microViz)
```

Atlas output files from version 2 are stored in the `MouseExample` folder.

```{r}
atlas_wd_folder <- "../MouseExample/"

taxonomy_file <- paste0(atlas_wd_folder, "genomes/taxonomy/gtdb_taxonomy.tsv")
tree_file <- paste0(atlas_wd_folder, "genomes/tree/gtdbtk.bac120.nwk")
quality_file <- paste0(atlas_wd_folder, "genomes/checkm/completeness.tsv")
counts_file <- paste0(atlas_wd_folder, "genomes/counts/raw_counts_genomes.tsv")
abundance_file <- paste0(atlas_wd_folder, "genomes/counts/median_coverage_genomes.tsv")
readstats_file <- paste0(atlas_wd_folder, "stats/read_counts.tsv")
keggmodules_file <- paste0(atlas_wd_folder, "genomes/annotations/dram/kegg_modules.tsv")
```


### Load metadata 
```{r}
metadata <- read.table(file.path(atlas_wd_folder, "metadata.txt"), sep = "\t", header = TRUE, row.names = "X")

metadata <- metadata[c("Diet", "Body_weight")]

head(metadata)

group_variable <- "Diet"
```         


We confirm that the mice on high-fat diet really put more weight on.

```{r plot_metadata, fig.height=10mm}
ggplot(metadata, aes(x = Diet, y = Body_weight)) +
  geom_boxplot() +
  geom_beeswarm(cex = 5) +
  ylim(10, 35) +
  theme_minimal()
```

# Load Atlas output

## Taxonomy

```{r, read_tax}
Tax <- read_tsv(taxonomy_file, show_col_types = FALSE)
kable(Tax)
```

```{r, lab_tax}
# create a short label for each species
Tax <- Tax %>%
  mutate(Label = ifelse(is.na(species) & is.na(genus), paste0(family, " ", user_genome), species)) %>%
  mutate(Label = ifelse(is.na(Label), paste0(genus, " ", user_genome), Label)) %>%
  mutate(species = Label, Label = NULL)
```

## Draw tree

```{r, read_newick}
Tree <- ggtree::read.tree(tree_file)
Tree <- full_join(Tree, dplyr::rename(Tax, label = user_genome), by = "label")
```

```{r, highlight_tree, fig.height = 10,fig.width=13}
library(phytools)

tree_plot <- ggtree(Tree, layout = "daylight") # +
# geom_tiplab(aes(label=Label))




phylum_counts <- table(Tax$phylum)
phylum_to_label <- names(phylum_counts)[phylum_counts > 1]

phylum_colors <- RColorBrewer::brewer.pal(length(phylum_to_label), "Paired")
names(phylum_colors) <- phylum_to_label



for (phy in phylum_to_label) {
  phy_tax <- Tax %>% filter(phylum == phy)

  phy_mrca <- findMRCA(Tree@phylo, phy_tax$user_genome)


  tree_plot <- tree_plot +
    geom_cladelabel(
      node = phy_mrca,
      label = phy,
      fontsize = 12,
      fill = phylum_colors[phy],
      color = phylum_colors[phy],
      align = TRUE,
      offset = 0,
      offset.text = .1,
      barsize = 5,
      angle = 0,
      hjust = 0,
    )
}



tree_plot
```

# Genome quality

```{r, read_genome_quality}
genome_quality <- read_tsv(quality_file, show_col_types = FALSE) %>%
  mutate(Quality_Score = Completeness - (5 * Contamination)) %>%
  left_join(Tax, by = c("Bin Id" = "user_genome")) %>%
  mutate(Name = species) %>%
  dplyr::select(-species)
```

```{r, plot_contam_vs_quality}
plt <- ggplot(genome_quality, aes(
  x = Contamination, y = Completeness, color = phylum,
  text = paste("Bin Id:", `Bin Id`, "\nName", Name, "\nQuality score:", Quality_Score)
)) +
  geom_point() +
  xlab("Contamination [%]") +
  ylab("Completeness [%]") +
  theme_minimal()
plotly::ggplotly(plt, tooltip = c("text"))
```

## Relative abundance

For the relative abundance, we take the coverage over the genome, not the raw counts.
This implicitly normalizes for genome size. The coverage is calculated as the median of the coverage values calculated in 1kb blocks.

```{r}
D <- read_tsv(abundance_file, show_col_types = FALSE) %>%
  column_to_rownames("...1") %>%
  as.matrix()
kable(topleft(D, c = 10))

# calculate relative abundance
rel_ab <- sweep(D, 1, rowSums(D), `/`)
```

```{r, relative_abund}
# get most abundant genomes
counts_per_genome <- data.frame(sums = colSums(rel_ab)) %>%
  rownames_to_column(var = "Genome") %>%
  left_join(Tax, by = c("Genome" = "user_genome")) %>%
  arrange(desc(sums))


ggplot(counts_per_genome %>%
  top_n(sums, n = 10), aes(x = reorder(species, -sums), y = sums)) +
  geom_col() +
  labs(x = "", y = "Abundance [rel_ab]", title = "Most abundant genomes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

### Phyloseq

Let's create a phyloseq object for further analysis.

```{r phloseq}
nw_tree <- ape::read.tree(tree_file)

# if you have real metadate the better

if (!exists("metadata")) {
  metadata <- data.frame(row.names = rownames(rel_ab))
  metadata$group <- "NoGroup"
}


pseq <- phyloseq(
  otu_table(rel_ab, taxa_are_rows = FALSE),
  tax_table(as.matrix(column_to_rownames(Tax, "user_genome"))),
  phy_tree(nw_tree),
  sample_data(metadata)
)

# fix taxonomy
pseq <- pseq %>%
  tax_fix()

microViz::phyloseq_validate(pseq)
```

### Ordination analysis

Here we use the recommended robust Centered log ratio transformation with PCA.
This corresponds to a PCoA with Aitchison distances.



```{r,fig.width=5,dpi=200}
ordination <- pseq %>%
  tax_transform(rank = "species", trans = "rclr") %>%
  ord_calc(
    method = "auto"
  )


ordination %>%
  ord_plot(
    plot_taxa = 1:3,
    colour = "Diet", fill = "Diet",
    size = 2
  ) +
  stat_ellipse(aes(linetype = Diet, colour = Diet), linewidth = 0.3)
```


### Interactive ordination

Run this command in the console
``` r

microViz::ord_explore(pseq)


```
click on stop sign on the toè right side to interrupt the interactive ordination.



### Typical bar chart

```{r}
pseq %>% comp_barplot(tax_level = "phylum", facet_by = "Diet") +
  coord_flip() +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```


Heatmap
```{r}
cols <- distinct_palette(n = 2, add = NA)
names(cols) <- unique(samdat_tbl(pseq)$Diet)




htmp <- pseq %>%
  tax_transform("compositional", rank = "genus") %>%
  tax_filter(min_prevalence = 0.25, prev_detection_threshold = 0.001, use_counts = FALSE) %>%
  comp_heatmap(
    grid_col = NA,
    cluster_rows = FALSE, row_title = NULL,
    # row_split = 5, #seq_len(ntaxa(ps_get(pseq))),
    tax_anno = taxAnnotation(
      Prev. = anno_tax_prev(bar_width = 0.9, size = grid::unit(1, "cm"), border = F)
    ),
    sample_anno = sampleAnnotation(
      # method one
      group = anno_sample("Diet"),
      col = list(group = cols), border = TRUE
    )
  )
htmp %>% ComplexHeatmap::draw(
  annotation_legend_list = attr(htmp, "AnnoLegends")
)
```

# Statistical models

Use linear models on CLR transformed data on all taxonimic levels. 

```{r}
lm_models <- pseq %>%
  tax_agg(rank = "species") %>%
  tax_prepend_ranks() %>%
  # Filtering
  tax_filter(min_prevalence = 0.1, undetected = 0, use_counts = FALSE) %>%
  tax_transform(
    trans = "clr", chain = TRUE, zero_replace = "halfmin", rank = "species"
  ) %>%
  taxatree_models(
    type = lm,
    ranks = NULL, # uses every rank available except the first
    variables = c(group_variable),
    # use_future =TRUE # paralellize
  )



# get stats and correct P values
lm_stats <- taxatree_models2stats(lm_models, conf.int = TRUE, conf.level = 0.95)
lm_stats <- taxatree_stats_p_adjust(
  data = lm_stats, method = "BH", grouping = "rank"
)

# lm_Stats is a phyloseq like object

lm_stats %>% taxatree_stats_get()
```

```{r, fig.width=5, fig.height=7}
plot_data <- lm_stats %>%
  taxatree_stats_get() %>%
  filter(rank == "species" & term == "DietHF") %>%
  dplyr::arrange(estimate)

ggplot(plot_data, aes(y = reorder(taxon, estimate), x = estimate)) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(
      xmin = conf.low,
      xmax = conf.high
    ),
    width = .2
  ) +
  geom_vline(xintercept = 0, linetype = "dashed")
```

```{r tax-tree-plot, fig.width=10, fig.height=5.5, dpi=300}
lm_stats %>%
  taxatree_label(
    rank == "phylum" | rank == "family",
    p.adj.BH.rank < 0.1
  ) %>%
  taxatree_plots(
    # node_size_range = c(1, 3), var_renamer = toupper,
    sig_stat = "p.adj.BH.rank",
    sig_threshold = 0.1,
  ) %>%
  .[[1]] %>% # show only the first plot
  taxatree_plot_labels(
    taxon_renamer = function(x) stringr::str_remove(x, "[pfgsL]: "),
    fun = ggrepel::geom_label_repel, x_nudge = 0.1, hjust = 0.5, size = 4
  )
```



# Functional annotations of genomes

The relative abundance of functional annotations per sample

The abundance is calculated as the sum of the relative abundance of all bacteria containing a function.

## Kegg modules produced by Dram

```{r, load_modules}
kegg_modules <- read_tsv(keggmodules_file, col_select = -1)
module_names <- kegg_modules %>%
  dplyr::select(c("module", "module_name")) %>%
  distinct() %>%
  column_to_rownames("module")


kable(head(kegg_modules))
```

```{r, module_presence}
step_coverage_threshold <- 0.8

module_step_coverage_matrix <- pivot_wider(kegg_modules,
  id_cols = genome,
  names_from = module,
  values_from = step_coverage
) %>%
  column_to_rownames("genome") %>%
  as.matrix()

module_step_coverage_matrix <- module_step_coverage_matrix[, colSums(module_step_coverage_matrix) > 0]

setHook("grid.newpage", function() pushViewport(viewport(x = 1, y = 1, width = 0.9, height = 0.9, name = "vp", just = c("right", "top"))), action = "prepend")
hatmap <- pheatmap(module_step_coverage_matrix, show_colnames = F, show_rownames = F)
setHook("grid.newpage", NULL, "replace")
grid.text("Modules", y = -0.07, gp = gpar(fontsize = 16))
grid.text("Genomes", x = -0.07, rot = 90, gp = gpar(fontsize = 16))



module_presence_matrix <- 1 * (module_step_coverage_matrix > step_coverage_threshold)

module_presence_matrix <- module_presence_matrix[, colSums(module_presence_matrix) > 0]

hatmap
```

```{r,heatmap_interactive2}
d_tax <- column_to_rownames(Tax, "user_genome")



annotations <- matrix("", nrow = nrow(module_step_coverage_matrix), ncol = ncol(module_step_coverage_matrix))
colnames(annotations) <- colnames(module_step_coverage_matrix)
rownames(annotations) <- rownames(module_step_coverage_matrix)

for (genome in rownames(annotations))
{
  for (module in colnames(annotations))
  {
    annotations[genome, module] <- paste0(
      "Name: ", d_tax[genome, "species"],
      "\nPhylum: ", d_tax[genome, "phylum"],
      "\nPathway: ", module_names[module, "name"]
    )
  }
}




heatmaply(module_step_coverage_matrix,
  custom_hovertext = annotations,
  row_side_colors = list(Phylum = d_tax[rownames(module_step_coverage_matrix), "phylum"]),
  showticklabels = c(FALSE, FALSE)
)
```

### Module abundance

Calculate Abundance of pathways as the sum of abundance of species where a module is presence. This is equal to the matrix multiplication.

```{r, module_abundance}
stopifnot(dim(rel_ab)[2] == dim(module_presence_matrix)[1])
module_rel_ab <- as.matrix(rel_ab) %*% module_presence_matrix


setHook("grid.newpage", function() pushViewport(viewport(x = 1, y = 1, width = 0.9, height = 0.9, name = "vp", just = c("right", "top"))), action = "prepend")
pheatmap(module_rel_ab, show_rownames = F, show_colnames = F)

setHook("grid.newpage", NULL, "replace")
grid.text("Modules", y = -0.07, gp = gpar(fontsize = 16))
grid.text("Samples", x = -0.07, rot = 90, gp = gpar(fontsize = 16))
```

```{r, heatmap-modules}
annotations <- matrix("", nrow = nrow(module_rel_ab), ncol = ncol(module_rel_ab))
colnames(annotations) <- colnames(module_rel_ab)
rownames(annotations) <- rownames(module_rel_ab)
for (sample in rownames(annotations))
{
  for (module in colnames(annotations))
  {
    annotations[sample, module] <- paste0(
      "Pathway: ", module_names[module, ]
    )
  }
}




heatmaply(module_rel_ab,
  custom_hovertext = annotations,
  showticklabels = c(FALSE, FALSE),
  row_side_colors = list(Diet = metadata[rownames(module_rel_ab), "Diet"]),
  colors = "RdBu"
)
```

```{r}
# get most abundant modules
abundance_per_module <- data.frame(abundance = colMeans(module_rel_ab)) %>% arrange(desc(abundance))
abundance_per_module <- cbind(abundance_per_module, module_names[rownames(abundance_per_module), ])
colnames(abundance_per_module) <- c("Average_abundance", "Description")






ggplot(abundance_per_module, aes(x = Average_abundance)) +
  geom_histogram() +
  labs(x = "Average module abundance", y = "counts") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```



Todo: statistical test on pathways...
